Addition and removal energies of circular quantum dots

### 2017-06-13 Manuscript/paper.tex

In this work we will mainly work within the formalism of second quantization; see for example Ref.\ \cite{shavitt2009many}.  To properly describe fermions, we introduce the annihilation and creation operators $\hat a_p$ and $\hat a_p^\dagger$. These operators must satisfy the anticommutation relations that encode the Pauli principle, namely
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p q}.
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.\ \eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.  In this paper, we use the letters $p, q, r, s, t, \ldots$ to label arbitrary single-particle \textit{states} in the basis.  A summation such as $\sum_p$ is then understood as a summation over every single-particle basis state $p$.  This is not to be confused with summations such as $\sum_j$ from \eqref{eq:onetwobodyhamiltonian}, which contains a label $j$ denoting an arbitrary \textit{particle}.  We thus see another benefit with second quantization: the Hamiltonian representation makes no attempt to distinguish particles at all, as expected for particles with quantum statistics.

@@

The goal then is to find the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, now given by
\begin{align}
  E_{\Phi} &= \sum_i n_i H^{[1] \prime}_{i i} + \sum_{i j} n_i n_j H^{[2] \prime}_{i j i j} \label{eq:hfenergy},
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{[1] \prime}_{p q} &\equiv \sum_{r s} C_{r p}^* \langle r | \hat{H}^{[1]} | s \rangle C_{s q}^{} \label{eq:hftransform1} \\
  H^{[2] \prime}_{p q r s} &\equiv \sum_{t u v w} C_{t p}^* C_{u q}^* \langle t u | \hat{H}^{[2]} | v w \rangle C_{v r}^{} C_{w s}^{}, \label{eq:hftransform2}
\end{align}
and $n_p$ is the \textit{occupancy}, defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state in $\Phi$}, \\
    0 & \text{if $p$ is an unoccupied state in $\Phi$}.
  \end{cases}
\end{align}
We will also use $\bar n_p \equiv 1 - n_p$ as a shorthand for the complement.

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree--Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm{F} \bm{C} = \bm{C} \bm{\varepsilon},
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p q} \equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_{r s i} n_i C_{r i}^* \langle p r | \hat{H}^{[2]} | q s \rangle C_{s i}^{},
\end{align}
and $\bm{\varepsilon}$ is a vector of Lagrange multipliers: each element $\varepsilon_{q'}$ is associated with a specific single-particle state $|q'\rangle$.

@@

The second-quantized operators in Eq.\ \eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1 i_2 \ldots i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators.  The normal ordering of the quasiparticle field operators is straightforward to compute.  What is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $|\Phi\rangle$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &=
  \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p q}^{}.
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_i n_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} n_i n_j \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_i n_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix $F_{p q}$.  The operator $\hat H$ in Eq.\ \eqref{eq:normordhamiltonian} is completely equivalent to $\hat H$ in Eq.\ \eqref{eq:vacuumhamiltonian}.  The only difference is that the meaning of a ``$k$-body operator'' has been redefined with respect to the reference state $\Phi$ rather than vacuum state $\ket{}$, causing matrix elements to be reshuffled among the $k$-body operators.  Where this makes a critical difference is when the operator expressions are \emph{truncated}, i.e.\ some $k$-body operators may be discarded from the computation for efficiency reasons.  Normal-ordering can drastically alter the importance of the $k$-body operators with respect to each other.

@@

By expanding the commutator in Eq.\ \eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H) \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \antisymm_{A B} D^{[k]}(A, B), \label{eq:commutmxe} \\
  \antisymm_{x y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr),
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} n_i \bar{n}_a A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} n_i n_j \bar{n}_a \bar{n}_b A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p q}(A, B)
  &\equiv
    \sum_r A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} n_i n_j \bar{n}_a A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} n_i \bar{n}_a \bar{n}_b A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} n_i \bar{n}_a \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right),
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    4 \antisymm_{p q} \antisymm_{r s}
    \sum_{i a} n_i \bar{n}_a A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    \notag \\
  &\hphantom{=}
    + \frac{1}{2} \sum_{i j} n_i n_j A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} \bar{n}_a \bar{n}_b A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + 2 \sum_t \left(
    \antisymm_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \antisymm_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}
    \right).
    \label{eq:flow2}
\end{align}

@@

The White generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta^{\prime [1]}_{a i} &\equiv \frac{\bar{n}_a n_i H^{[1]}_{a i}}{\tilde{\Delta}^{[1]}_{a i}} &
\eta^{\prime [2]}_{a b i j} &\equiv \frac{\bar{n}_a \bar{n}_b n_i n_j H^{[2]}_{a b i j}}{\tilde{\Delta}^{[2]}_{a b i j}}
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}^{[1]}_{a i} &\equiv \Delta_{a i} - H^{[2]}_{a i a i}, \notag \\
\tilde{\Delta}^{[2]}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H^{[1]}_{p_i p_i} -  H^{[1]}_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H^{[2]}_{a b a b} - H^{[2]}_{a i a i} - H^{[2]}_{b i b i} \notag \\
  &\qquad + H^{[2]}_{i j i j} - H^{[2]}_{a j a j} - H^{[2]}_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

Coupled-cluster theory is based on expressing the $N$-particle correlated wave function $\ket{\Psi}$ using the exponential ansatz,
\begin{align*}
  \ket{\Psi} = e^{\hat{T}} \ket{\Phi},
\end{align*}
where $\ket{\Phi}$ is the reference state as before.  The cluster operator $\hat{T} \equiv \hat{T}_1 + \hat{T}_2 + \cdots + \hat{T}_N$, is composed of $k$-particle $k$-hole excitation operators, $\hat{T}_k$,
\begin{align} \label{eq:cc_amps}
  \hat{T}_k \equiv \left(\frac{1}{k!}\right)^2 \sum_{\substack{a_1 \ldots a_k \\ i_1 \ldots i_k}} \bar{n}_{a_1} \ldots \bar{n}_{a_k} n_{i_1} \ldots n_{i_k} t^{[k]}_{a_1 \ldots a_k i_1 \ldots i_k} \normord{\hat{a}_{a_1}^\dagger \ldots \hat{a}_{a_k}^\dagger \hat a_{i_k}^{} \ldots \hat{a}_{i_1}^{}},
\end{align}
where the unknown matrix elements, $t^{[k]}_{a_1 \ldots a_k i_1 \ldots i_k}$, are known as \textit{cluster amplitudes} \cite{shavitt2009many}.

@@

We can then express the perturbation expansion in terms of summations over matrix elements as we did for the IM-SRG flow equation, again with the aid of Wick's theorem or diagrammatic techniques.  This leads to the following expression for the second-order correction:
\begin{align*}
  \varepsilon_p^{(2)}
  &=
    \sum_{i a b} \frac{n_i \bar{n}_a \bar{n}_b |H^{[2]}_{p i a b}|^2}{2 \Delta_{p i a b}}
    - \sum_{i j a} \frac{n_i n_j \bar{n}_a |H^{[2]}_{i j p a}|^2}{2 \Delta_{i j p a}},
\end{align*}
where the M\o ller--Plesset denominators $\Delta$ are defined in Eq.\ \eqref{eq:moellerplessetdenominator}.  The above expression is depicted in Figure \ref{fig:diagrams-sfe}.  Since there are numerous terms in the third-order correction $\varepsilon_p^{(3)}$, they are listed in diagrammatic form in Figure \ref{fig:diagrams-sfe}.  We will refer to QDPT to third order as ``QDPT3''.

@@

Particle attached and particle removed equations of motion (EOM) methods can be coupled with either IM-SRG or coupled cluster calculations. The principal idea is that one may construct a ladder operator $\hat{X}$ that promotes the $N$-particle ground state to any state in the $N + 1$ or $N - 1$ spectrum,
\begin{equation}\label{eq:ladder_def}
  \ket{\Psi^{(N \pm 1)}_u}  = \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
where $\hat{X}$ is in principle a linear combination of excitation and de-excitation operators that change particle number by one,
\begin{align}
  \label{eq:gen_attached}
  \hat{X}^{(N + 1)}_u &= \sum_p \bar{n}_p x^{[1, 0]}_p  \hat{a}^\dagger_p + \frac{1}{2} \sum_{p q r} \bar{n}_p \bar{n}_q n_r x^{[2, 1]}_{p q r} \hat{a}^\dagger_p \hat{a}^\dagger_q \hat{a}_r^{} + \cdots,  \\
  \label{eq:gen_removed}
  \hat{X}^{(N - 1)}_u &= \sum_p n_p x_p^{[0, 1]} \hat{a}_p^{} + \frac{1}{2} \sum_{p q r} n_p n_q \bar{n}_r  x^{[1, 2]}_{p q r} \hat{a}^\dagger_r \hat{a}_q^{} \hat{a}_p^{}  + \cdots.
\end{align}
Substitution of Eq.\ \eqref{eq:ladder_def} into the energy eigenvalue problem
\begin{gather*}
  \hat H \ket{\Psi_u^{(N \pm 1)}} = E^{(N\pm1)}_u \ket{\Psi_u^{(N \pm 1)}},
\end{gather*}
gives
\begin{equation}\label{eq:EOM}
  [\hat{H}, \hat{X}^{(N \pm 1)}_u] \ket{\Psi^{(N)}_0} = \pm \varepsilon^{(\pm)}_u \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
which constitutes a generalized eigenvalue problem for the amplitudes $x$, where $\varepsilon^{(\pm)}_u$ are the single-particle addition ($+$) and removal ($-$) energies. The quality of this calculation depends on the ansatz for the $N$-particle ground state, as well as the systematically improvable truncation on the ladder operators. In this work we include 1p and 2p1h excitations in the $N + 1$ ladder operator and likewise 1h and 2h1p operators for the $N - 1$ ladder operators.

@@

Despite these gains, the EOM calculation is still a partial diagonalization method, limited by the truncation to 2-particle-1-hole and 1-particle-2-hole operators. We expect $N + 1$ (or $N - 1$) states to be described appropriately by EOM-IM-SRG if their wavefunctions are dominated by 1-particle-0-hole (or 0-particle-1-hole) contributions in the rotated frame. We use partial norms of the EOM ladder operators to make this determination:
\begin{align}
  \label{eq:partial_norms_p}
  n_{\text{1-particle}} &= \sqrt{\sum_p \bar{n}_p | \bar{x}^{[1,0]}_p |^2},\\
  \label{eq:partial_norms_h}
  n_{\text{1-hole}} &= \sqrt{\sum_p n_p | \bar{x}^{[0,1]}_p |^2}.
\end{align}
Large single particle partial norms indicate that the EOM truncation is valid for the relevant state. States with lower single particle norms should be treated with a higher EOM approximation, which can be accomplished directly or perturbatively \cite{PhysRevC.95.044304}.

>>>>

In this paper, we use the letters $p, q, r, s, \ldots$ to label arbitrary \textit{single-particle states} in the basis.  In the discussion of many-body methods that rely a single, distinguished Slater determinant $\Phi$, known as the \textit{reference state}, we use the letters $i, j, k, l, \ldots$ as placeholders for the occupied states of $\Phi$, and $a, b, c, d, \ldots$ for the unoccupied states.  A summation such as $\sum_p$ is understood as a summation over every single-particle basis state $p$.  In contrast, $\sum_i$ only sums over those that are occupied and $\sum_a$ only sums over those that are unoccupied.

We will describe many-body methods within the formalism of second quantization; see for example Ref.\ \cite{shavitt2009many}.  To properly describe fermions, we introduce the annihilation and creation operators $\hat a_p$ and $\hat a_p^\dagger$. These operators must satisfy the anticommutation relations that encode the Pauli principle, namely
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p q}.
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.\ \eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.

@@

The goal then is to find the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, now given by
\begin{align}
  E_{\Phi} &= \sum_i H^{[1] \prime}_{i i} + \sum_{i j} H^{[2] \prime}_{i j i j} \label{eq:hfenergy},
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{[1] \prime}_{p q} &\equiv \sum_{r s} C_{r p}^* \langle r | \hat{H}^{[1]} | s \rangle C_{s q}^{} \label{eq:hftransform1} \\
  H^{[2] \prime}_{p q r s} &\equiv \sum_{t u v w} C_{t p}^* C_{u q}^* \langle t u | \hat{H}^{[2]} | v w \rangle C_{v r}^{} C_{w s}^{}, \label{eq:hftransform2}
\end{align}

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree--Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm{F} \bm{C} = \bm{C} \bm{\varepsilon},
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p q} \equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_{r s i} C_{r i}^* \langle p r | \hat{H}^{[2]} | q s \rangle C_{s i}^{},
\end{align}
and $\bm{\varepsilon}$ is a vector of Lagrange multipliers: each element $\varepsilon_{q'}$ is associated with a specific single-particle state $|q'\rangle$.

@@

The second-quantized operators in Eq.\ \eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1 i_2 \ldots i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators.  The normal ordering of the quasiparticle field operators is straightforward to compute.  What is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $|\Phi\rangle$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &= \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p q}^{}.
\end{align*}
and $n_p$ is the \textit{occupancy} of the single-particle state $p$, defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state of $\Phi$}, \\
    0 & \text{if $p$ is an unoccupied state of $\Phi$}.
  \end{cases}
\end{align}

The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix $F_{p q}$.  The operator $\hat H$ in Eq.\ \eqref{eq:normordhamiltonian} is completely equivalent to $\hat H$ in Eq.\ \eqref{eq:vacuumhamiltonian}.  The only difference is that the meaning of a ``$k$-body operator'' has been redefined with respect to the reference state $\Phi$ rather than vacuum state $\ket{}$, causing matrix elements to be reshuffled among the $k$-body operators.  Where this makes a critical difference is when the operator expressions are \emph{truncated}, i.e.\ some $k$-body operators may be discarded from the computation for efficiency reasons.  Normal-ordering can drastically alter the importance of the $k$-body operators with respect to each other.

@@

By expanding the commutator in Eq.\ \eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H) \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \antisymm_{A B} D^{[k]}(A, B), \label{eq:commutmxe} \\
  \antisymm_{x y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr),
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p q}(A, B)
  &\equiv
    \sum_r A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right),
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    4 \antisymm_{p q} \antisymm_{r s} \sum_{i a} A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    + \frac{1}{2} \sum_{i j} A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + 2 \sum_t \left(
    \antisymm_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \antisymm_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}
    \right).
    \label{eq:flow2}
\end{align}

@@

The White generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta^{\prime [1]}_{a i} &\equiv \frac{H^{[1]}_{a i}}{\tilde{\Delta}^{[1]}_{a i}} &
\eta^{\prime [2]}_{a b i j} &\equiv \frac{H^{[2]}_{a b i j}}{\tilde{\Delta}^{[2]}_{a b i j}}
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}^{[1]}_{a i} &\equiv \Delta_{a i} - H^{[2]}_{a i a i}, \notag \\
\tilde{\Delta}^{[2]}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H^{[1]}_{p_i p_i} -  H^{[1]}_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H^{[2]}_{a b a b} - H^{[2]}_{a i a i} - H^{[2]}_{b i b i} \notag \\
  &\qquad + H^{[2]}_{i j i j} - H^{[2]}_{a j a j} - H^{[2]}_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

Coupled-cluster theory is based on expressing the $N$-particle correlated wave function $\ket{\Psi}$ using the exponential ansatz,
\begin{align*}
  \ket{\Psi} = e^{\hat{T}} \ket{\Phi},
\end{align*}
where $\ket{\Phi}$ is the reference state as before.  The cluster operator $\hat{T} \equiv \hat{T}_1 + \hat{T}_2 + \cdots + \hat{T}_N$, is composed of $k$-particle $k$-hole excitation operators, $\hat{T}_k$,
\begin{align} \label{eq:cc_amps}
  \hat{T}_k \equiv \left(\frac{1}{k!}\right)^2 \sum_{\substack{a_1 \ldots a_k \\ i_1 \ldots i_k}} t^{[k]}_{a_1 \ldots a_k i_1 \ldots i_k} \normord{\hat{a}_{a_1}^\dagger \ldots \hat{a}_{a_k}^\dagger \hat a_{i_k}^{} \ldots \hat{a}_{i_1}^{}},
\end{align}
where the unknown matrix elements, $t^{[k]}_{a_1 \ldots a_k i_1 \ldots i_k}$, are known as \textit{cluster amplitudes} \cite{shavitt2009many}.

@@

We can then express the perturbation expansion in terms of summations over matrix elements as we did for the IM-SRG flow equation, again with the aid of Wick's theorem or diagrammatic techniques.  This leads to the following expression for the second-order correction:
\begin{align*}
  \varepsilon_p^{(2)}
  &=
    \sum_{i a b} \frac{|H^{[2]}_{p i a b}|^2}{2 \Delta_{p i a b}}
    - \sum_{i j a} \frac{|H^{[2]}_{i j p a}|^2}{2 \Delta_{i j p a}},
\end{align*}
where the M\o ller--Plesset denominators $\Delta$ are defined in Eq.\ \eqref{eq:moellerplessetdenominator}.  The above expression is depicted in Figure \ref{fig:diagrams-sfe}.  Since there are numerous terms in the third-order correction $\varepsilon_p^{(3)}$, they are listed in diagrammatic form in Figure \ref{fig:diagrams-sfe}.  We will refer to QDPT to third order as ``QDPT3''.

@@

Particle attached and particle removed equations of motion (EOM) methods can be coupled with either IM-SRG or coupled cluster calculations. The principal idea is that one may construct a ladder operator $\hat{X}$ that promotes the $N$-particle ground state to any state in the $N + 1$ or $N - 1$ spectrum,
\begin{equation}\label{eq:ladder_def}
  \ket{\Psi^{(N \pm 1)}_u}  = \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
where $\hat{X}$ is in principle a linear combination of excitation and de-excitation operators that change particle number by one,
\begin{align}
  \label{eq:gen_attached}
  \hat{X}^{(N + 1)}_u &= \sum_a x^{[1, 0]}_a  \hat{a}^\dagger_a + \frac{1}{2} \sum_{a b i} x^{[2, 1]}_{a b i} \hat{a}^\dagger_a \hat{a}^\dagger_b \hat{a}_i^{} + \cdots,  \\
  \label{eq:gen_removed}
  \hat{X}^{(N - 1)}_u &= \sum_i x_i^{[0, 1]} \hat{a}_i^{} + \frac{1}{2} \sum_{i j a} x^{[1, 2]}_{i j a} \hat{a}^\dagger_i \hat{a}_j^{} \hat{a}_a^{}  + \cdots.
\end{align}
Substitution of Eq.\ \eqref{eq:ladder_def} into the energy eigenvalue problem
\begin{gather*}
  \hat H \ket{\Psi_u^{(N \pm 1)}} = E^{(N\pm1)}_u \ket{\Psi_u^{(N \pm 1)}},
\end{gather*}
gives
\begin{equation}\label{eq:EOM}
  [\hat{H}, \hat{X}^{(N \pm 1)}_u] \ket{\Psi^{(N)}_0} = \pm \varepsilon^{(\pm)}_u \hat{X}^{(N \pm 1)}_u \ket{\Psi^{(N)}_0},
\end{equation}
which constitutes a generalized eigenvalue problem for the amplitudes $x$, where $\varepsilon^{(\pm)}_u$ are the single-particle addition ($+$) and removal ($-$) energies. The quality of this calculation depends on the ansatz for the $N$-particle ground state, as well as the systematically improvable truncation on the ladder operators. In this work we include 1p and 2p1h excitations in the $N + 1$ ladder operator and likewise 1h and 2h1p operators for the $N - 1$ ladder operators.

@@

Despite these gains, the EOM calculation is still a partial diagonalization method, limited by the truncation to 2-particle-1-hole and 1-particle-2-hole operators. We expect $N + 1$ (or $N - 1$) states to be described appropriately by EOM-IM-SRG if their wavefunctions are dominated by 1-particle-0-hole (or 0-particle-1-hole) contributions in the rotated frame. We use partial norms of the EOM ladder operators to make this determination:
\begin{align}
  \label{eq:partial_norms_p}
  n_{\text{1-particle}} &= \sqrt{\sum_a | \bar{x}^{[1,0]}_a |^2},\\
  \label{eq:partial_norms_h}
  n_{\text{1-hole}} &= \sqrt{\sum_i | \bar{x}^{[0,1]}_i |^2}.
\end{align}
Large single particle partial norms indicate that the EOM truncation is valid for the relevant state. States with lower single particle norms should be treated with a higher EOM approximation, which can be accomplished directly or perturbatively \cite{PhysRevC.95.044304}.
